### 2014-07-31 pot/LaTeX/skewnormal.tex

\section{Introduction}\label{s:intro}

@@

We first specify a model for the complete data, $Y_t(\bs)$, and then study the induced model for thresholded data, $\Yt_t(\bs)$.  
The full data model is given in Section \ref{s:model} assuming a skew normal distribution with a different variance each day.
Computationally, the values below the threshold are updated using standard Bayesian missing data methods as described in Section \ref{s:comp}.
The skew normal representation is from \citep{Minozzo2012} and is the sum of a normal and half-normal random variable.
\subsection{Half-normal distribution}\label{s:hn}
Let $u = |z|$ where $Z \sim N(\mu, \sigma^2)$.
Specifically, we consider the case where $\mu = 0$. Then $U$ follows a half-normal distribution which we denote as $U \sim HN(0, 1)$, and the density is given by 
\begin{align}
  f_U(u) = \frac{ \sqrt{2} }{ \sqrt{\pi \sigma^2} } \exp \left( - \frac{ u^2 }{ 2 \sigma^2 } \right) I(u > 0)
\end{align}
When $\mu = 0$, the half-normal distribution is also equivalent to a $N_{(0, \infty)}(0, \sigma^2)$ where $N_{(a, b)}(\mu, \sigma^2)$ represents a normal distribution with mean $\mu$ and standard deviation $\sigma$ that has been truncated below at $a$ and above at $b$.

\subsection{Complete data}\label{s:model}
Consider a skew Gaussian spatial process
\begin{align} \label{eq:fullmodel}
  Y_t(\bs) &= X_t(\bs) \beta + \sigma_1 z_t(\bs) + v_t(\bs)
\end{align}
where $z_t(\bs) = z_{tl}$ if $s \in P_{tl}$ where $P_{t1}, \ldots, P_{tL}$ form a partition, and $z_{tl} \iid N_{(0, \infty)}(0, 1)$, $\sigma_1 \in \calR$, $\sigma_2, \sigma_0 \in \calR^+$, and $v_t(\bs)$ is a spatial Gaussian process with mean zero and variance $\sigma_{tl}^2$.
It can be shown \citep{Zhang2010} that $Y_t(\bs)$ follows a skew normal distribution with skewness parameter $\alpha = \frac{ \sigma_1 }{ \sigma_{tl} }$.
We can then reexpress the model in (\ref{eq:fullmodel}) as
\begin{align}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha z_t(\bs) + v_t(\bs)
\end{align}
where $z_t(\bs) = z_{tl}$, $z_{tl} \iid N_{(0, \infty)}(0, \sigma_{tl}^2)$, and $v_t(\bs)$ is defined as before.

We model this with a Bayesian hierarchical model as follows.
Let $w_{t1}, \ldots, w_{tL}$ be partition centers so that
\begin{align*}
  P_{tl} = \{\bs_t : l = \argmin_k || \bs_t - w_{tk} ||\}.
\end{align*}
Then
\begin{align}
    Y_t(\bs) \mid \Theta, z_{t1}, \ldots, z_{tL} &= X_t(\bs) \beta + \alpha z_t(\bs) + v_t(\bs) \label{eq:hier}\\
    z_{tl}(\bs) \mid \Theta &\sim N_{(0, \infty)}(0, \sigma^2_{tl})\\
    v_t(\bs) \mid \Theta &\sim \Matern(0, \Sigma)\\
    \sigma^2_{tl} &\iid IG(\alpha, \beta)\\
    \alpha &\sim N(0, 10)\\
    w_{tk} &\sim Unif(\calD)
\end{align}
where $\Theta = \{w_{t1}, \ldots, w_{tL}, \beta, \sigma_t, \delta, \rho, \nu \}$; $l = \argmin_k ||\bs - w_k ||$; $\Sigma_t$ is a \Matern covariance matrix with variance $\sigma_{tl}^2 (1 - \delta^2)$, spatial range $\rho$ and smoothness $\nu$; and $\calD$ is the spatial domain of interest.

\section{Computation}\label{s:comp}
The MCMC for this model is fairly straightforward.
First, we impute values below the threshold.
Then, we update $\Theta$ using random walk MH or Gibbs sampling when appropriate.
Finally, we make spatial predictions.
Each requires the joint distribution for the complete data given $\Theta$.
As defined in \ref{eq:hier}, the distribution of $Y_t(\bs) \mid \Theta$ is the usual multivariate normal distribution with a \Matern spatial covariance structure.

\subsection{Imputation}\label{s:impute}
We can use Gibbs sampling to update $\Yt_t(\bs)$ for observations that are below $T$, the thresholded value. Given $\Theta$, $Y_t(\bs)$ has truncated normal full conditional with these parameter values.
So we sample $Y_t(\bs) \sim N_{(-\infty, T)}$

>>>>

\section{Introduction}\label{s:intro}
In most climatological applications, researchers are interested in learning about the average behavior of different climate variables (e.g. ozone, temperature, ranfall).
However, averages do not help regulators prepare for the unusual events that only happen once every 100 years.
For example, it is important to have an idea of how much rain will come in a 100-year floor in order to construct strong enough river levees to protect lands from flooding.
Unlike multivariate normal distributions, it is challenging to model multivariate extreme value distributions (e.g. generalized extreme value and generalized Pareto distribution) because few closed-form expressions exist for the density in more than two-dimensions \citep{Coles1991}.

In a bivariate setting, the extremal coefficient describes the pairwise dependence between two spatial locations \citep{Smith1990}.
Consider a spatial process $Y(\bs) \in \calR^n$ observed at locations $s \in \calD \subset \calR^2$.
Then the extremal coefficient, $\theta(\bs_i, \bs_j) \in [1, 2]$, is defined as
\begin{align}
  \Pr(Y(\bs_i) < c, Y(\bs_j) < c) = \Pr(Y(\bs_i) < c)^{\theta(\bs_i, \bs_j) }. \label{eq:extremalcoef}
\end{align}
One way to characterize the dependence over the entire set of spatial locations is to calculate all of the pairwise extremal coefficients.
Although this method provides information regarding the spatial structure of the observations, it does not fully characterize the joint spatial dependence.
One way to account for this limitation is the use of pairwise composite likelihoods \citep{Padoan2010,Blanchet2011,Huser2013}.

Recently, skew eliptical distributions have become a popular alternative to model correlated extreme values \citep{Genton2004,Zhang2010,Padoan2011}.
Specifically, the skew-normal and skew-t distribution offer a flexible way to handle non-symmetric data within a framework of multivariate normal and multivariate t-distributions.
In this paper, we present a modified version of the multivariate skew-t distribution that incorporates thresholding and random spatial partitions.
The advantage of using a thresholded model as opposed to a non-thresholded model is that is allows for the tails of the distribution to inform the predictions in the tails \citep{DuMouchel1983}.
The random spatial partitions increase the flexibility of the model by giving us a way to account for observations in spatial regions that may have higher variability due to certain regional climate variables (e.g. forest fires).

@@

We first specify a model for the complete data, $Y_t(\bs)$, and then study the induced model for thresholded data, $\Yt_t(\bs)$.
The full data model is given in Section \ref{s:data} assuming a skew normal distribution with a different variance each day.
Computationally, the values below the threshold are updated using standard Bayesian missing data methods as described in Section \ref{s:comp}.
The skew normal representation is from \citep{Zhang2010}.
\subsection{Half-normal distribution}\label{s:hn}
Let $u = |z|$ where $Z \sim N(\mu, \sigma^2)$.
Specifically, we consider the case where $\mu = 0$. Then $U$ follows a half-normal distribution which we denote as $U \sim HN(0, 1)$, and the density is given by
\begin{align}
  f_U(u) = \frac{ \sqrt{2} }{ \sqrt{\pi \sigma^2} } \exp \left( - \frac{ u^2 }{ 2 \sigma^2 } \right) I(u > 0)
\end{align}
When $\mu = 0$, the half-normal distribution is also equivalent to a $N_{(0, \infty)}(0, \sigma^2)$ where $N_{(a, b)}(\mu, \sigma^2)$ represents a normal distribution with mean $\mu$ and standard deviation $\sigma$ that has been truncated below at $a$ and above at $b$.

\subsection{Complete data}\label{s:data}
Consider a skew Gaussian spatial process
\begin{align} \label{eq:fullmodel}
  Y_t(\bs) &= X_t(\bs) \beta + \sigma_1 z_t(\bs) + v_t(\bs)
\end{align}
where $z_t(\bs) = z_{tl}$ if $s \in P_{tl}$ where $P_{t1}, \ldots, P_{tL}$ form a partition, and $z_{tl} \iid N_{(0, \infty)}(0, 1)$, $\sigma_1 \in \calR$, and $v_t(\bs)$ is a spatial Gaussian process with mean zero and variance $\sigma_{tl}^2$.
It can be shown \citep{Zhang2010} that $Y_t(\bs)$ follows a skew normal distribution with skewness parameter $\alpha = \frac{ \sigma_1 }{ \sigma_{tl} }$.
We can then reexpress the model in (\ref{eq:fullmodel}) as
\begin{align}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha z_t(\bs) + v_t(\bs)
\end{align}
where $z_t(\bs) = z_{tl}$, $z_{tl} \iid N_{(0, \infty)}(0, \sigma_{tl}^2)$, and $v_t(\bs)$ is defined as before.

We model this with a Bayesian hierarchical model as follows.
Let $w_{t1}, \ldots, w_{tL}$ be partition centers so that
\begin{align*}
  P_{tl} = \{\bs_t : l = \argmin_k || \bs_t - w_{tk} ||\}.
\end{align*}
Then
\begin{align}
    Y_t(\bs) \mid \Theta, z_{t1}, \ldots, z_{tL} &= X_t(\bs) \beta + \alpha z_t(\bs) + v_t(\bs) \label{eq:hier}\\
    z_{tl}(\bs) \mid \Theta &\sim N_{(0, \infty)}(0, \sigma^2_{tl})\\
    v_t(\bs) \mid \Theta &\sim \Matern(0, \Sigma_t(\lambda))\\
    \sigma^2_{tl} &\iid IG(\alpha, \beta)\\
    \alpha &\sim N(0, 10)\\
    w_{tk} &\sim Unif(\calD)
\end{align}
where $\Theta = \{w_{t1}, \ldots, w_{tL}, \beta, \sigma_t, \alpha, \lambda, \rho, \nu \}$; $l = \argmin_k ||\bs - w_k ||$; $\Sigma_t(\lambda)$ is a \Matern covariance matrix with variance $\sigma_{tl}^2$, spatial range $\rho$, smoothness $\nu$, and $\lambda$ is the percentage of variation that can be attributed to the partial sill; and $\calD$ is the spatial domain of interest.

\section{Computation}\label{s:comp}
The MCMC for this model is fairly straightforward.
First, we impute values below the threshold.
Then, we update $\Theta$ using random walk MH or Gibbs sampling when appropriate.
Finally, we make spatial precictions using conditional multivariate normal results and the fact that the distribution of $Y_t(\bs) \mid \Theta, z_{tl}$ is the usual multivariate normal distribution with a \Matern spatial covariance structure.

\subsection{Imputation}\label{s:impute}
We can use Gibbs sampling to update $Y_t(\bs)$ for observations that are below $T$, the thresholded value. Given $\Theta$, $Y_t(\bs)$ has truncated normal full conditional with these parameter values.
So we sample $Y_t(\bs) \sim N_{(-\infty, T)}$
